---
title: "rEDM Primer"
author: "Hao Ye"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
---

# Abstract

This document is designed to be a quick intro for "empiricial dynamic modeling" (EDM) and the **rEDM** package. For more detailed information, please see the `rEDM-tutorial` vignette include in the package [[link]](https://cran.r-project.org/web/packages/rEDM/vignettes/rEDM-tutorial.html).

# Theory

Suppose that we have a system with $d$ state variables: $x_1, x_2, \dots, x_d$. We can represent the state at a specific time, $t$, as a vector with these $d$ components: $\bar{x}(t) = \langle x_1(t), x_2(t), \dots, x_d(t) \rangle$.

We generally assume that the system is deterministic, such that the current state fully determines future states:
$$\bar{x}(t+1) = F \left(\bar{x}(t)\right)$$
though we may not know what this function $F$ actually is.

## Example

In model systems, we usually know what $F$ is, since we have differential or difference equations that describe the system behavior. For example, the coupled logistic map equations are:
$$\begin{align*}
x_1(t+1) &= r_1 x_1(t) \left[1 - x_1(t) - \alpha_{1,2} x_2(t)\right]\\
x_2(t+1) &= r_2 x_2(t) \left[1 - x_2(t) - \alpha_{2,1} x_1(t)\right]
\end{align*}$$
with parameters $r_1, r_2, \alpha_{1,2}, \alpha_{2,1}$.

## Time Delay Embedding (Delay Coordinate Embedding)

We would naturally expect that to predict the future of any component of the system, e.g. $x_i$, we would need to know all the other $x_j$ values. Takens' Theorem actually posits that this is not strictly necessary, and that under certain conditions, a single time series is sufficient.

In other words, instead of relying on:
$$x_i(t+1) = F_i\left(x_1(t), x_2(t), \dots, x_d(t)\right)$$

the system dynamics can be represented as a function of a single variable and its lags:
$$x_i(t+1) = G_i\left(x_i(t), x_i(t-1), \dots, x_i(t-(E-1))\right)$$

Note that we use $E$ coordinates instead of $d$ and that we do not make any claims about how the functions $F_i$ and $G_i$ are related.

## Usage

Since we generally don't know the form of $G_i$, we stick to inferring it from the data. The **rEDM** package provides several methods to do so, but here we just stick to simplex projection.

Simplex projection uses a weighted nearest-neighbors approximation to estimate $G_i$. That's kind of a fancy way of saying the following:

1. Suppose we have the value of $x$ and its lags at time $s$. Then we want a prediction of $x(s+1) = G\left(x(s), x(s-1), \dots, x(s - (E-1))\right)$.

2. We look for $j = 1..k$ nearest neighbors in the observed time series of $x$ such that $\langle x(s), x(s-1), \dots, x(s - (E-1))\rangle \approx \langle x(n_j), x(n_j-1), \dots, x(n_j - (E-1))\rangle$.

3. We then suppose that $x(s+1) \approx x(n_j+1)$.

Mathematically, this occurs by using some distance function to judge how similar $\langle x(s), x(s-1), \dots, x(s - (E-1))\rangle$ is to $\langle x(n_j), x(n_j-1), \dots, x(n_j - (E-1))\rangle$ and estimating $x(s+1)$ as a weighted average of the $x(n_j+1)$ values with weighting determined by the distances.

# Applications

These principles of EDM can be used for several different applications. These include:

* univariate forecasting (predicting future behavior using lags of a single time series)
* multivariate forecasting (predicting future behavior using lags of multiple time series)
* characterizing various properties of time series (e.g. predictability and "nonlinearity")
* identifying causal relationships between time series

Here, we'll just look at univariate forecasting, though there are examples for the others among the package documentation. (see the vignettes on [CRAN](https://cran.r-project.org/web/packages/rEDM/index.html))

# Univariate Forecasting

## Data

We'll start by looking at sunspot data. Note that there is also monthly data, but it seems quite a bit noisier:
```{r}
dat <- data.frame(yr = as.numeric(time(sunspot.year)), 
                  sunspot_count = as.numeric(sunspot.year))

plot(dat$yr, dat$sunspot_count, type = "l")
```

## Determining E

From the above sections, you may recall that one of the parameters is the embedding dimension, `E`. Since we ultimately want to test forecast skill, we use only the first 2/3 of the data to fit `E`. There are several other parameters that we can set, but the defaults work fine for our case:

```{r}
library(rEDM)                    # load the package

n <- NROW(dat)
lib <- c(1, floor(2/3 * n))      # indices for the first 2/3 of the time series
pred <- c(floor(2/3 * n) + 1, n) # indices for the final 1/3 of the time series

simplex(dat,                     # input data (for data.frames, uses 2nd column)
        lib = lib, pred = lib,   # which portions of the data to train and predict
        E = 1:10)                # embedding dimensions to try
```

The output is a data.frame describing how well the model "predicted" on the given data. The columns are various statistical outputs. We'll focus just on the measures of model performance:

* rho (correlation between observed and predicted values)
* mae (mean absolute error)
* rmse (root mean squared error)

It looks like `E = 3` or `4` is optimal. Since we generally want a simpler model, if possible, let's go with `E = 3` to forecast the remaining 1/3 of the data.

## Forecasts

We can use similar code to produce forecasts for the last 1/3 of the time series. Since we want the outputs in addition to the performance statistics, we add an additional argument to the function call:

```{r}
output <- simplex(dat,
        lib = lib, pred = pred,  # predict on last 1/3
        E = 3, 
        stats_only = FALSE)      # return predictions, too
```

Here the output is again a data.frame with a list column for the predictions:

```{r}
predictions <- output$model_output[[1]]
str(predictions)
```

Let's plot the predictions against the original data:
```{r}
plot(dat$yr, dat$sunspot_count, type = "l")
lines(predictions$time, predictions$pred, col = "blue", lty = 2)
```

Since we also have an estimate of the prediction uncertainty, let's include that as well:
```{r}
plot(dat$yr, dat$sunspot_count, type = "l")
lines(predictions$time, predictions$pred, col = "blue", lty = 2)
polygon(c(predictions$time, rev(predictions$time)), 
        c(predictions$pred - sqrt(predictions$pred_var), 
        rev(predictions$pred + sqrt(predictions$pred_var))), 
        col = rgb(0, 0, 1, 0.5), 
        border = NA)
```

