
```{r}
library(forecast)
```


## Data setup

```{r}
data = read.csv("portal_timeseries.csv", stringsAsFactors = FALSE)
head(data)
NDVI_ts = ts(data$NDVI, start = c(1992, 3), end = c(2014, 11), frequency = 12)
rain_ts = ts(data$rain, start = c(1992, 3), end = c(2014, 11), frequency = 12)
```

## Plot the data

```{r}
plot(NDVI_ts)

```

## White noise model

* Most basic time-series model 
* Normally distributed data w/fixed mean and variance
* No change time-series is just random samples

> y = c + e_t, where e_t ~ N(0, sigma)

```{r}
avg_model = meanf(NDVI_ts)
plot(NDVI_ts)
lines(fitted(avg_model), col = 'red')
summary(avg_model)
```


## Auto-regressive model

* But we know this isn't right

```{r}
acf(NDVI_ts)
pacf(NDVI_ts)
```

* Build a model that takes this autocorrelation into account
* "Autoregressive model""
* Predict value based on previous states in time-series

> y_t = c + b1 * y_t-1 + b2 * y_t-2 ... + e_t

* Does this remind you of a biological model?
    * This model is bascially a Gompertz population model if y is log(N)

* ARIMA: autoregressive, integrated, moving average
* Fit using `Arima()` function

```{r}
arima_model = Arima(NDVI_ts, c(2, 0, 0))
plot(NDVI_ts)
lines(fitted(arima_model), col = 'red')
summary(arima_model)
```


* Check fit

```{r}
plot(resid(arima_model))
acf(resid(arima_model))
```

* Autocorrelation at 1 and 2 years
* Seasonal signal
* Seasonal component is modeled in the same way, but in 1 full season steps

> y_t = constant + b1 * y_t-1 + b2 * yt_2 + b3 * y_t-12 + b4 * y_t-24 + e_t

```{r}
seasonal_arima_model = Arima(NDVI_ts, c(2, 0, 0), seasonal = c(2, 0, 0))
plot(NDVI_ts)
lines(fitted(seasonal_arima_model), col = 'red')
summary(seasonal_arima_model)
acf(resid(seasonal_arima_model))
```


> CLASS EXERCISE: Build seasonal and non-seasonal ARIMAs for the precipitation data


## Automating fits

* Iterative manual fitting is great, but
    * Often want "best" model
    * Fit lots of models at once
* Fit many possible values of seasonal, AR, differencing, and MA, and pick the best
* Unit root tests, minimization of the AICc and MLE

```{r}
arima_model = auto.arima(NDVI_ts)
plot(NDVI_ts)
lines(fitted(arima_model), col='red')
summary(arima_model)
```

* MA component in best model
* What are those other two numbers
    * Moving average: autocorrelated errors
    * Differencing: handles strong one-step autocorrelations, like trends

* Can use `seasonal = FALSE` to skip seasonal signal

## Incorporating external co-variates

* Often want external predictors in our model
* Do this like we would in normal regression
* Add values of x and associated coefficient to model

> y_t = c + b1 * y_t-1 + b2 * x_t ... + e_t

```{r}
rain_arima_model = auto.arima(NDVI_ts, xreg = rain_ts)
plot(NDVI_ts)
lines(fitted(rain_arima_model), col = 'blue')
lines(fitted(arima_model), col = 'red')
```

